Uniform formula lower bounds for testing the existence of and counting satisfiable assignments

### 2022-06-27 article/main.tex
Report is done

\documentclass[a4paper, 11pt]{article}
% \usepackage{ae,lmodern}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[USenglish]{babel}
\usepackage[margin=3cm]{geometry}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{todonotes}
\usepackage{enumitem}
% \usepackage{algorithm}
% \usepackage[noend]{algpseudocode}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section] % Number within sec ?
% \newtheorem{theorem}{Theorem}
\newtheorem{prop}[theorem]{Proposition}
\newtheorem{property}[theorem]{Property}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{observation}[theorem]{Observation}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
% \newtheorem*{remark*}{Remark}
% \theoremstyle{definition}
% \newtheorem{problem}{Problem}
% \newtheorem{openproblem}{Open Problem}

@@

\newcommand{\todohere}{\todo[inline]{TODO}}%
\newcommand{\todoh}[1]{\todo[inline]{TODO: #1 }}%

@@

\todoh{Motivation: understanding complexity. Explicit lower bounds: programs, circuits, formulas.
Uniformity.}

A line of work initiated by Fortnow~\cite{fortnow2000time,fortnow2005time} in 1997 
has proven model-independent time-space lower bounds for the SAT
problem and related problems within the polynomial-time hierarchy.
This approach was then extended by Williams~\cite{williams2006inductive,williams2007time,williams2013alternation} 
into a complex framework which culminated in proving that \SAT{} 
does not have a small-space algorithm that runs in time $n^{2\cos(\pi/7)}  \simeq n^{1.801}$.
This line of work concluded~\cite{buss2015limits} with a proof 
that the $2\cos(\pi/7)$ exponent is optimal for this framework,
which implies that radically new ideas would be required to separate \NP{} from \Log{}.
This approach was later extended to obtain slightly stronger lower bounds
for quantum classes, and against small-space randomized computation~\cite{mudigonda2020time}.

@@

Finding strong lower bounds on the formula size of explicit functions in \NP{}
is a long-standing open problem.
The best lower bounds known depend on the computational basis, that is, 
the set of gates that are allowed in the formula.
For the full binary basis $B_2$, Nechiporuk\cite{nechiporuk66boolean}
proved in 1966 an $\Omegat{n^2}$ lower bound for an explicit function in \NP{}, 
which remains unbeaten to this day.
For the De Morgan basis $\{\wedge, \vee, \neg\}$, Andreev~\cite{andreev1987method} 
proved an $\Omegat{n^{2.5}}$ lower bound, 
which was later improved by H{\aa}stad~\cite{hastad1998shrinkage} to $\Omegat{n^{3}}$.
This lower bound is optimal up to polylog factors,
since their function has (uniform) formulas of size $\Ot{n^{3}}$.

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Results overview}

The \NAND{} basis is rather expressive: Andreev's function has \NAND{} formulas of depth $(3+o(1)) \log n$ (and therefore of size $\Ot{n^3}$), which matches the lower bound for the De Morgan Basis.
\todoh{more about nand?}

Namely, we prove the following:
\begin{theorem}\label{thm:main}
	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT{} does not have 
	uniform \NAND{} formulas of depth $(c + o(1)) \log n$.
\end{theorem}

@@

\subsection{Uniform formulas}
A boolean formula $\phi$ over $n$ inputs is a 
rooted full binary tree\footnote{\textit{Full} means that every node has either two or zero children.}
whose internal nodes, called gates, are labeled by a function $f: \bit^2 \rightarrow \bit$,
and leaves are labeled with a variable, $x_i$ its negation $\neg x_i$ for some $i, 1\le i \le n$,
or a constant $c\in\bit$.
The depth of a formula is the depth of the underlying tree.
Given an input $x \in\bit^n$, the value of $\phi$ at $x$ can be defined 
as the value of it root on input $x$, 
where the value of a gate is defined inductively as follows:
\[val(g) =\begin{cases}
	c & \text{ if $g$ is labeled with the constant} c\in\bit\\
	x_i & \text{ if $g$ is labeled with $x_i$}\\
	\neg x_i & \text{ if $g$ is labeled with $\neg x_i$}\\
	f(g_1, g_2) & \text{ if $g$ is labeled  $f$, and $g_1,g_2$ are children of $g$}
\end{cases}\]

@@

A family of formulas is a sequence $(\phi_n)_{n\in\NN}$ of formulas such that
for every i, $\phi_i$ is a formula over $i$ inputs.
A language $L\subseteq \bit^*$
is computed by\footnote{In what follows, we may use ``$L$ has formulas ...'' for ``$L$ is computed by a family of formulas ...''} 
the family of formulas $(\phi_n)_{n\in\NN}$
if for every $n$ and every $x\in\bit^n$, $\phi_n(x) = 1 \Leftrightarrow x\in L$.

@@

\begin{lemma}[Speedup rule]\label{lemma:speedup}
	For every $d > 0$ and $0 \ge x \ge d$, we have
	\begin{flalign*}
	\NDL{d} &\subseteq (\exists n^{x/2}) (\forall n) \NDL{d-x}\text{ and}\\
	\NDL{d} &\subseteq (\forall n^{x/2}) (\exists n) \NDL{d-x}
	\end{flalign*}
\end{lemma}

The intuitive idea behind this lemma is that the only
information needed to know that an OR gate evaluates
to 1 is that one of its children evaluates to 1. 
We apply this idea to every OR gate in the first $x \log n$
levels of the formula,
and then use a universal quantifier to handle the remaining AND gates.

The role of the existential quantifier is to guess for every OR gate in the first $x\log n$ levels,
which of its children evaluates to 1, and we only evaluate the corresponding subformula.
Moreover, making this guess reduces the number of OR gates that we will evaluate
(since the formula alternate between levels of OR and AND gates, this number doubles every two levels,
instead of doubling at every level), 
therefore we only need to guess $O(n^{x/2})$ bits.

We can then eliminate OR gates and replace them with their selected child.
We are left with a formula starting with $(x/2)\log n$ levels of AND gates, 
followed by smaller OR/AND formulas of depth $(d-x + o(1)) \log n$.
The main formula evaluates to 1 if and only if for every of these $O(n^{x/2})$
smaller formulas evaluates to true: 
we can check this using a universal quantifier, 
which requires $O(\frac{x}{2}\log n)$ bits, and then running an $\NDL{d-x}$ computation.
Note that while the second quantifier only requires $O(\frac{x}{2}\log n)$ nondeterministic bits,
it still computes and outputs $n^{1+o(1)}$ bits, hence the $n^1$ exponent.

@@

\begin{proof}[Proof of Lemma~\ref{lemma:speedup}]
	We prove the result for $\exists\forall$ computations, 
	the result for $\forall\exists$ computations follows from the fact
	that $\NDL{d}$ is closed under taking the complement.

	Let $L\in\NDL{d}$, let $M : (x,i,n) \mapsto g$ be 
	a descriptor machine for $L$, and let $x$ be an input of length $n$.

	We now construct a machine $M'$ that describes $\NDL{d-x}$-formulas,
	such that an $\exists\forall$ computation with $M'$ as its last stage accepts $L$.
	On an input $x$

	We describe how to build bit strings $y$ of length $n^{x/2}$ and
	\todoh{finish details of proof}
\end{proof}

Lemma~\ref{lemma:speedup} yields the following corollary,
which gives us a simulation of formulas by Turing machines that is more efficient
than the naive evaluation.
\begin{corollary}\label{cor:speedup}
	For every $d > 0$, we have
	\[\NDL{d} \subseteq \NTIME[n^{d/2}] \cap \coNTIME[n^{d/2}].\]
\end{corollary}
\begin{proof}
	Applying Lemma~\ref{lemma:speedup} with $x = d$, we obtain
	$\NDL{d} \subseteq (\exists n^{d/2}) (\forall n) \NDL{0}$.
	However, looking closely at the proof of Lemma~\ref{lemma:speedup}
	reveals that the $\forall$ quantifier here only uses $O(\frac{d}{2}\log n)$
	nondeterministic bits.
	Therefore, we can use exhaustive search over the $n^{d/2}$ bit strings of length $\frac{d}{2}\log n$
	to remove that quantifier.
	This leaves us we a deterministic $n^{o(1)}$ computation, and we get that
	$\NDL{d} \subseteq (\exists n^{d/2}) \TIME[n^{d/2}] = \NTIME[n^{d/2}]$.

@@

We now prove the Conditional Depth hierarchy theorem, 
stated without proof in Section~\ref{sec:overview}.
\begin{theorem}
	If there exists $c > 0$ such that $\SAT \subseteq \NDL{c}$, 
	then for all $b > a > 0$, we have
	\[\NDL{a} \subsetneq \NDL{b}.\] 
\end{theorem}
\begin{proof}

\end{proof}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Structure of optimal proofs}
In this section, we describe
the process that helps us build a proof, formed of successive applications
of the speedup and slowdown rules, 
that reaches a contradiction from the assumption
$\SAT\in\NDL{c}$ for some $c < 4\cos(\pi /7)$.

@@

Eq.~\ref{eq:cstr_ak} leads us to the following definition.
\begin{definition}
	Let $c > 2\phi$.
	A sequence of real number $(a_k)_{k=0,\ldots,t}$ is \textit{well-behaved for $c$}\todo{find a name for that}
	if the following holds:
	\begin{enumerate}
		\item for every $k=0,\ldots,t, a_k \ge 1$,
		\item $a_t = 1$,
		\item for every $k=0,\ldots,t, c a_{k+1}/2  \ge a_k$, and
		\item\label{constr:decr} for every $k=0,\ldots,t-1, a_{k+1} < \frac{4a_k}{c(c-2)}$.
	\end{enumerate}
\end{definition}
Notice that constraint~\ref{constr:decr} above implies that $a_{k+1} < a_k$,
as $c > 2\phi$ implies that $\frac{4}{c(c-2)} < 1$.

@@

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Can we turn depth lower bounds into (good) size lower bounds?}
Can we prove an improved Brent/Spira-type theorem that turns our lower bounds into super-cubic size lower bounds?
Useful starting point: Size-Depth Tradeoffs for Boolean Formulae (Bonet, Buss).

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion and future work}
The lower bounds given in this work are against a very restricted
type of formulas (NAND formulas, or equivalently alternating OR/AND formulas).
An open problem is to lift these lower bounds to other computational bases,
for example by allowing XOR and equality gates,
using ``even'' or ``odd-parity'' quantifiers, 
and assuming that \ParSAT{} has low-depth formulas.
In that case, we can get a lower bound for \SAT{} \textit{or} \ParSAT{}.
We can go further, and get lower bounds for \SharpSAT{},
since \ParSAT{} is the least significant bit of \SharpSAT{}.
The first idea to solve \SAT{} (that is, to remove existential quantifiers) 
with \SharpSAT{} would be to take the OR of all $n+1$
bits of \SharpSAT{}, but this add depth $\log n$ to the formula.
However, \textsf{Tautology} can be solved
with the most significant bit of \SharpSAT{}:
therefore we may be able to remove existential quantifiers 
by looking at a single bit of \SharpSAT{}
for a formula that corresponds to a negation of the verifier.

@@

\subsection{Overview of the approach}
Improves over Williams (cite): no need to enumerate proof annotations.

Therein, Williams enumerates all possible proof skeletons of length $k$,
and tries to find the best set of parameters for each of them, using a linear program.
In this work, we propose a different approach, that proves to be more efficient in practice:
we use an Integer Linear program whose solution is
an optimal proof of length $k$
along with its parameters.

More precisely, we give an ILP $P_{c,k}$ which is feasible if and only if
there exists a proof annotation $p$ of length $k$,\footnote{Our program takes as input a value $k'$ and searches over proofs with $k'$ application of the speedup rule, which implies $k'+1$ slowdowns since every quantifier must be removed to reach a contradiction. Therefore, we have $k = 2k'+1$.} along with parameters $x, r_0$
such that $p$ proves a contradiction under the assumption that $\SAT\in\NDL{c}$
Moreover, its solution encodes $p,q,r_0$.
By using binary search over $c$, we can quickly get an approximation 
of $c_k^*$, the largest $c$ such that a proof of length $k$ can reach a contradiction assuming that $\SAT\in\NDL{c}$.
Repeating the above steps for increasing values of $k$ shows that the sequence $(c_k^*)_{k}$
seems to converge to some value $c_\infty^* > 3.5$ (see Table~\ref{table:ilp_res}).

\begin{table}[htbp]
	\begin{center}
	\begin{tabular}{ |c|l|l| } 
	\hline
	$k$ & $c_k^*$ & Skeleton of $p_k^*$ \\ 
	\hline
	3 & 2.82812 & \texttt{100} \\
	7 & 3.20020 & \texttt{1100100} \\
	11 & 3.32617 & \texttt{11100100100} \\
	15 & 3.39844 & \texttt{111001010010100} \\
	19 & 3.44629 & \texttt{1111001001010010100} \\
	23 & 3.47559 & \texttt{11110010100101001010100} \\
	27 & 3.49805 & \texttt{111110010010100101001010100} \\
	\hline
	\end{tabular}
	\end{center}
	\caption{Approximate value of $c_k^*$ and skeleton of $p_k^*$ for some values of $k$.}\label{table:ilp_res}
\end{table}

>>>>

\documentclass[a4paper, 11pt]{article}
% \usepackage{ae,lmodern}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[USenglish]{babel}
\usepackage[margin=3cm]{geometry}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{todonotes}
\usepackage{enumitem}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section] % Number within sec ?
\newtheorem{prop}[theorem]{Proposition}
\newtheorem{property}[theorem]{Property}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{observation}[theorem]{Observation}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

@@

@@

As a first step to understand the relations between different complexity classes,
an important problem in complexity theory is to find
lower bounds lower bounds for explicit problems in various model: programs, circuits, or formulas.

A particular attention was dedicated to the \SAT{} problem, 
since a super-polynomial lower bound for this problem would imply $\NP \neq \P$
(for programs, $\NP \neq \textsf{P/poly}$ for circuits).

A line of work initiated by Fortnow~\cite{fortnow2000time,fortnow2005time} in 1997 
has proven model-independent time-space lower bounds
(or equivalently, lower bounds against small-space algorithms) for \SAT{}
and related problems within the polynomial-time hierarchy.
This approach was then extended by Williams~\cite{williams2006inductive,williams2007time,williams2013alternation} 
into a complex framework which culminated in proving that \SAT{} 
does not have a small-space algorithm that runs in time $n^{2\cos(\pi/7)}  \simeq n^{1.801}$.
This line of work concluded~\cite{buss2015limits} with a proof 
that the $2\cos(\pi/7)$ exponent is optimal for this framework,
which implies that radically new ideas would be required to separate \NP{} from \Log{}.
This approach was later extended to obtain slightly stronger lower bounds
for quantum classes, and against small-space randomized computation~\cite{mudigonda2020time}.

@@

Finding strong lower bounds on the formula size of explicit functions in \NP{}
is a long-standing open problem.
The best lower bounds known depend on the computational basis, that is, 
the set of gates that are allowed in the formula.
For the full binary basis $B_2$, Nechiporuk~\cite{nechiporuk66boolean}
proved in 1966 an $\Omegat{n^2}$ lower bound for an explicit function in \NP{}, 
which remains unbeaten to this day.
For the De Morgan basis $\{\wedge, \vee, \neg\}$, Andreev~\cite{andreev1987method} 
proved an~$\Omegat{n^{2.5}}$ lower bound, 
which was later improved by H{\aa}stad~\cite{hastad1998shrinkage} to $\Omegat{n^{3}}$.
This lower bound is optimal up to polylog factors,
since this function has (uniform) formulas of size $\Ot{n^{3}}$.

@@

%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Results overview}

In this work, we focus in the \NAND{} basis, 
that is formulas where every gate computes the function $\NAND: x,y\mapsto \neg(x\wedge y)$;
there are no internal, only inputs can be negated.
The \NAND{} basis is universal (i.e. can compute any boolean function), 
and is rather expressive: Andreev's function has \NAND{} formulas of depth~$(3+o(1)) \log n$
(and therefore of size $\Ot{n^3}$), which matches the lower bound for the De Morgan Basis.

Our main result is the following:
\begin{theorem}\label{thm:main}
	For any $c < 4 \cos(\pi/7) \simeq 3.603$, \SAT{} does not have 
	uniform \NAND{} formulas of depth $(c + o(1)) \log n$.
\end{theorem}

It is based on a novel algorithm that computes the 
value of a \NAND{} formula in $\Sigma_2$ sublinear time.
Our algorithm uses the equivalence between \NAND{} formulas and alternating OR/AND formulas.

@@

\subsection{Uniform formulas}
A boolean formula $\phi$ over $n$ inputs is a 
rooted full binary tree\footnote{\textit{Full} means that every node has either two or zero children.}
whose internal nodes, called gates, are labeled by a function $f: \bit^2 \rightarrow \bit$,
and leaves are labeled with a variable~$x_i$, its negation $\neg x_i$ for some $i, 1\le i \le n$,
or a constant $c\in\bit$.
The depth of a formula is the depth of the underlying tree.
Given an input $x \in\bit^n$, the value of $\phi$ at $x$ can be defined 
as the value of it root on input $x$, 
where the value of a gate is defined inductively as follows:
\[val(g) =\begin{cases}
	c & \text{ if $g$ is labeled with the constant} c\in\bit\\
	x_i & \text{ if $g$ is labeled with $x_i$}\\
	\neg x_i & \text{ if $g$ is labeled with $\neg x_i$}\\
	f(g_1, g_2) & \text{ if $g$ is labeled  $f$, and $g_1,g_2$ are children of $g$}
\end{cases}\]

@@

A family of formulas is a sequence $(\phi_n)_{n\in\NN}$ of formulas such that
for every i, $\phi_i$ is a formula over $i$ inputs.
A language $L\subseteq \bit^*$
is computed by\footnote{In what follows, we may use ``$L$ has formulas ...'' for ``$L$ is computed by a family of formulas ...''} 
the family of formulas~$(\phi_n)_{n\in\NN}$
if for every $n$ and every $x\in\bit^n$, $\phi_n(x) = 1 \Leftrightarrow x\in L$.

@@

\begin{lemma}[Speedup rule]\label{lemma:speedup}
	For every $d > 0$ and $0 \ge x \ge d$, we have
	\begin{flalign*}
	\NDL{d} &\subseteq (\exists n^{x/2}) (\forall n) \NDL{d-x}\text{ and}\\
	\NDL{d} &\subseteq (\forall n^{x/2}) (\exists n) \NDL{d-x}
	\end{flalign*}
\end{lemma}
\begin{proof}
	The intuitive idea behind this lemma is that the only
	information needed to know that an OR gate evaluates
	to 1 is that one of its children evaluates to 1. 
	We apply this idea to every OR gate in the first $x \log n$
	levels of the formula,
	and then use a universal quantifier to handle the remaining AND gates.

	The role of the existential quantifier is to guess for every OR gate in the first $x\log n$ levels,
	which of its children evaluates to 1, and we only evaluate the corresponding subformula.
	Moreover, making this guess reduces the number of OR gates that we will evaluate
	(since the formula alternate between levels of OR and AND gates, this number doubles every two levels,
	instead of doubling at every level), 
	therefore we only need to guess $O(n^{x/2})$ bits. 
	Let $y$ be the corresponding bitstring.

	We can then eliminate OR gates and replace them with their selected child.
	We are left with a formula starting with $(x/2)\log n$ levels of AND gates, 
	followed by smaller OR/AND formulas of depth $(d-x + o(1)) \log n$.
	The main formula evaluates to 1 if and only if for every of these $O(n^{x/2})$
	smaller formulas evaluates to true: 
	we can check this using a universal quantifier, 
	which requires $O(\frac{x}{2}\log n)$ bits, and then running an $\NDL{d-x}$ computation.
	Note that while the second quantifier only requires $O(\frac{x}{2}\log n)$ nondeterministic bits,
	it still computes and outputs $n + x\log n$ bits 
	(the output along with 1 bit for each of the first $x\log n$ levels), hence the $n^1$ exponent. 
	Let $z$ be the corresponding bitstring.

@@

	Consider the machine $M'$ that,
	on input $a'$ of length $n' = n + x\log n$ for some $n$,
	let $a' = ab$, where $a$ has length $n$ and $b$ has length $x\log n$,
	and $M'$ returns $M'(a', i, n') = M(a, bi, n)$.
	By construction, $M'$ describes $\NDL{d-x}$-formulas.
	Moreover, if the original formula was true on $x$, 
	then there exists a choice of $y$ such that for all $z$,
	the formula described by $M'$ is true on input $xb(y,z)$,
	where $b(y,z)$ is the bitstring that describes a path 
	encoded by $y$ restricted to the choices of $z$.
	This is an $\exists\forall$ computation with $M'$ as its last stage that accepts $L$.

	The result for $\forall\exists$ follows since $\NDL{d}$ is closed under complement.
\end{proof}

Lemma~\ref{lemma:speedup} yields the following corollary,
which gives us a simulation of formulas by Turing machines that is more efficient
than the naive evaluation.
\begin{corollary}\label{cor:speedup}
	For every $d \ge 1$, we have
	\[\NDL{d} \subseteq \NTIME[n^{d/2}] \cap \coNTIME[n^{d/2}].\]
\end{corollary}
\begin{proof}
	Applying Lemma~\ref{lemma:speedup} with $x = d$, we obtain
	$\NDL{d} \subseteq (\exists n^{d/2}) (\forall n) \NDL{0}$.
	However, looking closely at the proof of Lemma~\ref{lemma:speedup}
	reveals that the $\forall$ quantifier here only uses $O(\frac{d}{2}\log n)$
	nondeterministic bits.
	Therefore, we can use exhaustive search over the $n^{d/2}$ bit strings of length $\frac{d}{2}\log n$
	to remove that quantifier.
	This leaves us we a deterministic $n^{o(1)}$ computation, and we get that
	$\NDL{d} \subseteq (\exists n^{d/2}) \TIME[n^{d/2}] = \NTIME[n^{d/2}]$.

@@

We now prove the Conditional Depth hierarchy theorem, 
stated without proof in Section~\ref{sec:overview}.
\begin{theorem}
	If there exists $c > 0$ such that $\NTIME[n] \subseteq \NDL{c}$, 
	then for all $b > a > 0$, we have
	\[\NDL{a} \subsetneq \NDL{b}.\] 
\end{theorem}
\begin{proof}
	We give a proof by contradiction. Assume that there exists $0 < a < b$
	such that $\NDL{b} \subseteq \NDL{a}$. 
	A standard amplification by padding argument shows that this implies
	$\NDL{d} \subseteq \NDL{a}$ for \textit{any} $d$.
	More precisely, we show that $\NDL{r\cdot a} \subseteq \NDL{a}$
	implies  $\NDL{r^2\cdot a}\subseteq \NDL{r\cdot a} \subseteq \NDL{a}$ for any $r > 1$.
	Taking $r = b/a > 1$ and applying this result $t = \log_2\log_r d/a$ times, we get.
	$\NDL{r^{2^t}} = \NDL{d} \subseteq \NDL{a}$.

	Now, assuming that there exists $c \ge 1/2$ such that $\NTIME[n] \subseteq \NDL{c}$,
	we have:
	\begin{flalign*}
		\NTIME[2a] 
			&\subseteq \NDL{2ca}\\
			&\subseteq \NDL{a}\\
			&\subseteq \NTIME[a]
	\end{flalign*}
	The first inclusion follows from a padding argument, 
	the second one follows from the amplification argument above as $2c > 1$,
	the last inclusion uses efficient simulation of $\ND$ by RAMs.
	The proof for $c < 1/2$ is similar, except that the second inclusion follows
	from $2c < 1$.
	But this proves $\NTIME[2a] \subseteq \NTIME[a]$ for some $a > 0$,
	which contradicts the Nondeterministic Time Hierarchy Theorem.
\end{proof}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Structure of asymptotically optimal proofs}
In this section, we describe
the process that helps us build an quantifier-trading proof, 
formed of successive applications
of the speedup and slowdown rules along with their parameters, 
that reaches a contradiction from the assumption
$\SAT\in\NDL{c}$ for some $c < 4\cos(\pi /7)$.
While these proofs might not be the shortest that yield a contradiction
for a given $c$, they present a very simple inductive structure that makes them easier to analyze.

We introduce here the \textit{proof annotation} notation of Williams~\cite{mudigonda2020time},
which succinctly describes a quantifier-trading proof: 
let $1_x$ denote an application of the speedup rule with parameter $x$, 
and let $0$ denote an application of the slowdown rule.
The proof annotation of a quantifier-trading proof is the string corresponding
to the concatenation of the representation of each of its steps.
For example, the proof annotation corresponding to 
the short quantifier-trading proof presented in Section~\ref{sec:overview}
is $1_200$.

@@

Eq.~\ref{eq:cstr_ak} leads us to the following definition.
\begin{definition}
	Let $c > 2\phi$.
	A sequence of real number $(a_k)_{k=0,\ldots,t}$ is \textit{well-behaved for $c$}
	if the following holds:
	\begin{enumerate}
		\item for every $k=0,\ldots,t, a_k \ge 1$,
		\item $a_t = 1$,
		\item for every $k=0,\ldots,t, c a_{k+1}/2  \ge a_k$, and
		\item\label{constr:decr} for every $k=0,\ldots,t-1, a_{k+1} < \frac{4a_k}{c(c-2)}$.
	\end{enumerate}
\end{definition}
Notice that constraint~\ref{constr:decr} above implies that $a_{k+1} < a_k$,
as $c > 2\phi$ implies that $\frac{4}{c(c-2)} < 1$.

@@

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion and future work}
As mentioned in the introduction, the lower bounds 
proved in this work are \textit{depth} lower bounds,
which are weaker than size lower bounds, since the latter directly imply the former.
However, there exists theorems that give a converse, 
but the depth is reduced by a constant factor
that depend on the type of formula.
This follows from work of Spira~\cite{spira1971time}, Brent~\cite{brent1974parallel},
and more recently Bonet and Buss~\cite{bonet1994size}.
For NAND formulas, the best known theorems have constants that are greater than $2$,
therefore using them does not improve over the naive $2\cos \pi/7$
that follow from TM lower bounds.
However, for most families of formulas, the constants in the best theorems
are not known to be tight\footnote{See Sergeev~\cite{sergeev2019relation} (in russian) for a recent exposition 
best upper and lower bounds known on the constants of multiple classes of formulas.}, 
therefore, it might be possible to improve them
and use our result to prove nontrivial size lower bounds,
and maybe super-cubic size lower bounds.

The lower bounds given in this work are against a very restricted
type of formulas (NAND formulas, or equivalently alternating OR/AND formulas).
An open problem is to lift these lower bounds to other computational bases,
for example by allowing XOR and equality gates,
using ``even'' or ``odd-parity'' quantifiers, 
and assuming that \ParSAT{} has low-depth formulas.
In that case, we can get a lower bound for \SAT{} \textit{or} \ParSAT{}.
We can go further, and get lower bounds for \SharpSAT{},
since \ParSAT{} is the least significant bit of \SharpSAT{}.
The first idea to solve \SAT{} (that is, to remove existential quantifiers) 
with \SharpSAT{} would be to take the OR of all $n+1$
bits of \SharpSAT{}, but this add depth $\log n$ to the formula.
However, \textsf{Tautology} can be solved
with the most significant bit of \SharpSAT{}:
therefore we may be able to remove existential quantifiers 
by looking at a single bit of \SharpSAT{}
for a formula that corresponds to a negation of the verifier.

@@

In this section, we describe how we extend and improve 
the approach of Williams~\cite{williams2006inductive, williams2007time} 
to programmatically find the proof annotations of a given $c$ that yield a contradiction
for the largest possible $c$.

\subsection{Overview of the approach}
We call a \textit{proof skeleton} a proof annotation stripped of its speedup parameters.
For example, the proof skeleton of the annotation $1_200$ is $100$.

In his work, Williams enumerates all possible proof skeletons of length $k$,
and tries to find the best set of parameters for each of them, using a linear program.
In this work, we propose a different approach, that proves to be more efficient in practice:
we use an Integer Linear Program (ILP) whose solution is
an optimal proof annotation of length $k$.

More precisely, we give an ILP $P_{c,k}$ which is feasible if and only if
there exists a proof annotation $p$ of length $k$,\footnote{Our program takes as input a value $k'$ and searches over proofs with $k'$ application of the speedup rule, which implies $k'+1$ slowdowns since every quantifier must be removed to reach a contradiction. Therefore, we have $k = 2k'+1$.} along with parameters $x, d_0$
such that $p$ proves a contradiction under the assumption that $\SAT\in\NDL{c}$
starting from the class $\NDL{d_0}$.
Moreover, its solution encodes $p,q,d_0$.
By using binary search over $c$, we can quickly get an approximation 
of $c_k^*$, the largest $c$ such that a proof of length $k$ 
can reach a contradiction assuming that $\SAT\in\NDL{c}$.
Repeating the above steps for increasing values of $k$ shows that the sequence $(c_k^*)_{k}$
seems to converge to some value $c_\infty^* > 3.5$ (see Table~\ref{table:ilp_res}).

\begin{table}[htbp]
	\begin{center}
	\begin{tabular}{ |c|l|l| } 
	\hline
	$k$ & $c_k^*$ & Skeleton of $p_k^*$ \\ 
	\hline
	3 & 2.82812 & 100 \\
	7 & 3.20020 & 1100100 \\
	11 & 3.32617 & 11100100100 \\
	15 & 3.41797 & 111001001100100 \\
	19 & 3.46094 & 1111001001001100100 \\
	23 & 3.49902 & 11110010011001001100100 \\
	27 & 3.52930 & 111100011000111001001100100 \\
	\hline
	\end{tabular}
	\end{center}
	\caption{Approximate value of $c_k^*$ and skeleton of $p_k^*$ for some values of $k$.}\label{table:ilp_res}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Details of the model}
Given an integer $k \ge 3$ and a real $c$, we define the integer linear program $P_{c,k}$ as follows:
\[
\begin{array}{rrcll}
	(P_{k,c}) &\max& 1&&\\
	\text{Initial speedup constraints:}&&&&\\
	&d_1&=&d_0 - x_1&\\
	&a_{1, -1}&=&1&\\
	&a_{1, -2}&\ge&x_1/2&\\
	&su_1&=&1&\\
	\text{Speedup constraints:}&&&&\\
	\forall i \ge 1, su_i = 1\Rightarrow &d_i&=&d_{i-1} - x_i&\\
	\forall i \ge 1, su_i = 1\Rightarrow &a_{i, j}&\ge&a_{i-1, j+1}& j = 1,\ldots,k-1\\
	\forall i \ge 1, su_i = 1\Rightarrow &a_{i, -2}&\ge&x_i/2&\\
	\forall i \ge 1, su_i = 1\Rightarrow &a_{i, -1}&=&1&\\
	\text{Slowdown constraints:}&&&&\\
	\forall i \ge 1, su_i = 0\Rightarrow &d_i&\ge&c\cdot d_{i-1}/2&\\
	\forall i \ge 1, su_i = 0\Rightarrow &d_i&\ge&c\cdot a_{i-1, -1}&\\
	\forall i \ge 1, su_i = 0\Rightarrow &d_i&\ge&c\cdot a_{i-1, -2}&\\
	\forall i \ge 1, su_i = 0\Rightarrow &a_{i, j}&=&a_{i-1, j-1}& j = 2,\ldots,k\\
	\text{Well-formed proof constraints:}&&&&\\
	\forall i \ge 1, & 2\sum_{t=1}^i su_{t} &\ge& i&\\ 
					 & 2\sum_{i=1}^k su_{i} &=& k - 1&\\ 
	\text{Require contradiction:}&&&&\\
	& d_k &<&d_0&\\
	\text{Variables:}&&&&\\
	&d_i &\in& [0,+\infty), &i =0,\ldots, k\\
	&a_{i,j} &\in& [1,+\infty), &i,j =1,\ldots, k\\
	&x_i &\in& [0,+\infty), &i =1,\ldots, k\\
	&su_{i} &\in& \{0,1\}, &i =1,\ldots, k
\end{array}
\]
The variables have the following meaning:
\begin{enumerate}
	\item $d_i$: verifier is $\NDL{d_i}$ on line $i$,
	\item $a_{i,k+1-j}$: exponent of the $j$-th-to-last quantifier on line $i$. 
	We use $a_{i,-j}$ as a shorthand notation for $a_{i,k+1-j}$. 
	By indexing $a$ relative to the last quantifier instead of the first allows us to simplify
	the linear program: for example, 
	we do not need to keep track of the number of quantifiers on the current line. 
	\item $x_i$: if line $i$ is the result of a speedup, it is a speedup with parameter $x_i$,
	\item $su_{i}$: 1 if and only if line $i$ results in the application of a speedup, 0 otherwise.
\end{enumerate}
The constraints of the type ``$su_i = b$ implies $C$'' can be encoded using the so-called ``Big M method''\footnote{See \url{https://en.wikipedia.org/wiki/Big\_M\_method}.}. 
The constraint $d_k < d_0$ can be approximately encoded as $d_k \le d_0 - \delta$ for some small $\delta > 0$.
